Generalised gravitational burst generation with Generative Adversarial Networks

### 2020-11-20 Paper_draft/main.tex
add more text to CNN results section from Thurs meeting

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Generative Adversarial Networks}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% basic intro to GANs
%
A subset of deep learning that has seen fruitful development in recent years
are generative adversarial networks \acp{GAN}~\cite{Goodfellow2014}. These unsupervised algorithms learn patterns in a
given training data set using an adversarial process. The generations from
\acp{GAN} are currently state-of-the-art in fields such as high quality image
fidelity~\cite{brock2018large,karras2019analyzing}, text-to-image
translation~\cite{reed2016generative}, and video
prediction~\cite{liang2017dual} as well as time series
generations~\cite{esteban2017realvalued}.
%
% Basic components of a GAN
%
\acp{GAN} train two competing neural networks, consisting of a discriminator
network that is set up to distinguish between real and fake data and a
generator network that produces fake versions of the real data. The generator model performs a mapping from a fixed length vector $\mathbf{z}$ to its
representation of the data. The input vector is drawn randomly from a Gaussian distribution which is referred to as a latent space comprised of latent variables. The latent space is a compressed representation of a data distribution which the generator applies meaning to during training. Sampling points from this space allows the generator to produce a variety of different generations, with different points corresponding to different features in the generations. The discriminator maps its input $\mathbf{x}$ to a probability that the input came from either the training (real) data or
generator (fake).
\begin{comment}
Chris' definition of latent space if needed: It maps the intrinsic
variation of the training space onto the Gaussian latent space distribution.
\end{comment}
%
% training a GAN
%
During training, the discriminator and generator are updated using batches of data. Random latent vectors are given to the generator to produce a batch of fake samples and an equal batch of real samples is taken from the training data. The discriminator makes predictions on the real and fake samples and the model is updated through minimising the binary cross-entropy function \cite{Goodfellow-et-al-2016}
%
\begin{equation}
    L = -y \log(\hat{y}) - (1 - y) \log(1-\hat{y}),
    \label{eqn:crossentropy}
\end{equation}
where $\hat{y}$ is the network prediction and $y$ is the true output. While training the discriminator, D, on real data, we set $y = 1$ and $\hat{y} = D(\mathbf{x})$ which from \cref{eqn:crossentropy} gives $L(D(\mathbf{x}),1) = \log(D(\mathbf{x}))$. While training on fake data produced by the generator, G, $y = 0$ and $\hat{y} = D(G(\mathbf{z}))$ and so, $L(D(G(\mathbf{z})),0) = \log(1-(D(G(\mathbf{z}))))$. Since the objective of the discriminator is to correctly classify fake and real data these equations should be maximised, while the goal of the generator should be to minimize these equations. This gives the \ac{GAN} loss as
%
\begin{equation}
   \mathop{\text{min}}_{G}  \mathop{\text{max}}_{D} V(D,G) = \mathbb{E}_{\mathbf{x} \sim p_{\text{\text{data}}}(\mathbf{x})} [\text{log} D(\mathbf{x})] + \mathbb{E}_{\mathbf{z} \sim p_{\mathbf{z}}(\mathbf{z})} [\text{log}(1-D(G(\mathbf{z})))],
 \label{equation:GANloss}
 \end{equation}
where $p_{\text{\text{data}}}(\mathbf{x})$ is the distribution of real data and $p_{\text{z}}(\mathbf{z})$ is the distribution of the latent space. 
\begin{figure}[!h]
    \centering
    \includegraphics{figures/BCE-loss.png}
    \caption{Binary cross-entropy plotted as a function of possible predicted probabilities in the case where the true label $y$ is 1 for a sample size of 512.}
    \label{fig:BCE_loss}
\end{figure}
\subsection{Training stages}
%
Training a GAN is involves updating both the discriminator and generator in stages. First, the discriminator is updated on real instances from the training set. We set the true label $y=1$ and calculate the loss with respect to the predictions $\hat{y}$ via \cref{equation:GANloss}. Stochastic gradient descent is used to minimize the loss which has reduced to $L_D(\textrm{real}) = -\log(\hat{y})$. The discriminator is then trained on fake instances taken from the generator where we set $y=0$ and minimize $L_D(\textrm{fake}) = -\log(1-\hat{y})$. To train the generator, we use a composite model of the generator and discriminator and allow the gradients to flow through this entire model. At this step we set the labels for the generator as real $y=1$ and minimize $L_G = -\log(\hat{y})$. The act of switching the labels and minimizing this form of loss is identical to the previously mentioned generator loss but is beneficial as in practice this function provides more useful gradients. This tweak to the generator loss is called non-saturating generator loss and was reported in the original GAN paper \cite{Goodfellow2014}. It was also shown in that paper that if the generator and discriminator can no longer improve, then the discriminator can no longer distinguish between real and fake i.e. $D(x) = \frac{1}{2}$. In \cref{fig:BCE_loss} we show how the binary-cross entropy varies with probability inputs when the real label is set to 1. We can see that for optimal training of the discriminator of 0.5 the binary cross entropy sits around 0.7. One could then say that the optimal generator loss during training is 0.7 however in practice the generator loss may be higher due to the continually alternating training method.

@@

% describe the main idea
%
Given a 100-dimensional vector drawn from a normal distributed latent space and a one-hot encoded class
label, the GAN is able to generate burst-like waveforms generalised from the
training set. We set out by describing the quality of generated waveforms and
how they compare to the training set. We then explore the structure of the
latent and class spaces by interpolating between points in these spaces. We
test two methods of sampling from the class space that can be used to generate
a new signals composed of weighted elements of each training class.
\begin{figure}[!h]
    \centering
    \includegraphics[width=\textwidth]{figures/generations/sg.png}
    \includegraphics[width=\textwidth]{figures/generations/rd.png}
    \includegraphics[width=\textwidth]{figures/generations/wnb.png}
    \includegraphics[width=\textwidth]{figures/generations/blip.png}
    \includegraphics[width=\textwidth]{figures/generations/bbh.png}
    \caption{\ac{GAN} Generated waveforms plotted as a function of time. The latent space inputs for each panel is randomised and each row is assigned one of the five class vectors. By row: sine-Gaussian, Ringdown,
white noise burst, Gaussian pulse, binary black hole merger. On the first four rows each x axis is zoomed for ease of viewing, while the last row shows the full one second.}
\label{fig:gen_signals} 
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Known class signal generation}
In \cref{fig:gen_signals} we show conditional signal generations using our generator network. We can see the generations are of very good quality and appear like they could plausibly have come from the training set. We can also see that the model has learned the overall characteristics of the five training classes and is able to disentangle each class and associate them with the conditional input. Additionally, as the latent variable changes as do the parameters within each class. For instance \cref{fig:gen_signals} shows how signals vary in frequency, central epoch, decays and phase. The GANs ability to generate a variety of signals for various latent input indicate stable training and no mode collapse. 
\begin{figure}[!h]
    \centering
    \includegraphics[width=\textwidth]{figures/generations/z_interp_sg.png}
    \includegraphics[width=\textwidth]{figures/generations/z_interp_rd.png}
    \includegraphics[width=\textwidth]{figures/generations/z_interp_wnb.png}
    \includegraphics[width=\textwidth]{figures/generations/z_interp_blip.png}
    \includegraphics[width=\textwidth]{figures/generations/z_interp_bbh.png}
    \caption{Generated interpolated waveforms plotted as a function of time. For each interpolation two points were randomly chosen in the latent space and the between panels are linearly interpolated vectors between these two points. Each row keeps on of the class vetors constant throughout the latent space interpolation.}
    \label{fig:z_interp}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Interpolation within the latent space}
%
% expain what we are planning to show here
%
 We have shown that the generator produces quality signals and that the model responds well to randomly sampled Gaussian latent vectors. We now assume that during training the generator has learned a mapping from a Gaussian latent space to signal space and that this mapping is smooth. To test this, we fix the class vector input and linearly interpolate between two randomly chosen point in the latent space. In \cref{fig:z_interp} we show the generated waveforms, with the class vectors held constant along each row. We can see that each plot shows plausible waveforms suggesting that the generator has constructed a smooth traversable, space.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Interpolation between pairs of classes}

@@

\subsection{Results}
In~\cref{fig:eff_curves} we compare the \ac{CNN} results between the three
datasets, we train three \acp{CNN} on the vertex, simplex and uniform data sets
and use these models to make predictions on the other unseen datasets. We make
a comparison by fixing the fraction of samples incorrectly identified as
signals (false alarm rate) and plotting this versus the optimal SNR of the
signals.~\chris{this last sentence isn't quite right}

These results for a \ac{CNN} trained on vertex signals shows clear delineation from other two classes at snr$>$10. For the simplex trained network, the simplex dataset narrowly outperforms the other two sets and, finally, the uniform trained network shows similar performance on all datasets. The \ac{CNN} is
robust enough to capture the differences between the datasets and shows that
the GAN can generate a variety of unmodeled waveforms that can be used in
future testing.

@@

\begin{figure}[h!] % "[t!]" placement specifier just for this example
\begin{subfigure}{0.48\textwidth}
\includegraphics[width=\linewidth]{figures/conditional_trained.png}
\caption{}
\end{subfigure}\hspace*{\fill}
\begin{subfigure}{0.48\textwidth}
\includegraphics[width=\linewidth]{figures/efficiency_curve_vtest.png}
\caption{}
\end{subfigure}

\medskip
\begin{subfigure}{0.48\textwidth}
\includegraphics[width=\linewidth]{figures/simplex_trained.png}
\caption{}
\end{subfigure}\hspace*{\fill}
\begin{subfigure}{0.48\textwidth}
\includegraphics[width=\linewidth]{figures/efficiency_curve_stest.png}
\caption{}
\end{subfigure}

\medskip
\begin{subfigure}{0.48\textwidth}
\includegraphics[width=\linewidth]{figures/uniform_trained.png}
\caption{}
\end{subfigure}\hspace*{\fill}
\begin{subfigure}{0.48\textwidth}
\includegraphics[width=\linewidth]{figures/efficiency_curve_utest.png}
\caption{}
\end{subfigure}

\caption{Efficiency curves comparing the performance of the \acp{CNN}. The true alarm probability is plotted as a function of the optimal SNR of the signals. (a), (c), (e) show a CNN trained on
vertex generations (top), simplex generations (middle), uniform
generations (bottom) for a fixed false alarm rate of $10^{-3}$ and tested on all three data sets. (b), (d), (f) show fix the testing set vertex (top), simplex (middle), uniform (bottom) and show the performance of three trained CNNs on the different data sets.}  \jordan{This shows both comparisons of the models v test data and test data v models.} \label{fig:eff_curves}
\end{figure}

\begin{figure}[h!] % "[t!]" placement specifier just for this example
\begin{subfigure}{0.48\textwidth}
\includegraphics[width=\linewidth]{figures/efficiency_curve_vtest.png}
\caption{}
\end{subfigure}\hspace*{\fill}
\begin{subfigure}{0.48\textwidth}
\includegraphics[width=\linewidth]{figures/efficiency_curve_stest.png}
\caption{}
\end{subfigure}

\medskip
\begin{subfigure}{0.48\textwidth}
\includegraphics[width=\linewidth]{figures/efficiency_curve_utest.png}
\caption{}
\end{subfigure}\hspace*{\fill}
\begin{subfigure}{0.48\textwidth}
\includegraphics[width=\linewidth]{figures/efficiency_curve_mix_text.png}
\caption{}
\end{subfigure}

\caption{Efficiency curves comparing the performance of the \acp{CNN}. The true alarm probability is plotted as a function of the optimal SNR of the signals. Plots (a), (b), (c) fix the testing set vertex (top), simplex (middle), uniform (bottom) and show the performance of three trained CNNs on the different data sets for a fixed false alarm rate of $10^{-3}$. Plot (d) shows CNNs trained on vertex, an even mixture of vertex and simplex and an even mixture of vertex and uniform and tested against the vertex dataset}  \jordan{This shows just the trained models v a test set with additional "mixture plot"} \label{fig:eff_curves}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Generative Adversarial Networks}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% basic intro to GANs
%
A subset of deep learning that has seen fruitful development in recent years
are generative adversarial networks \acp{GAN}~\cite{Goodfellow2014}. These unsupervised algorithms learn patterns in a
given training data set using an adversarial process. The generations from
\acp{GAN} are currently state-of-the-art in fields such as high quality image
fidelity~\cite{brock2018large,karras2019analyzing}, text-to-image
translation~\cite{reed2016generative}, and video
prediction~\cite{liang2017dual} as well as time series
generations~\cite{esteban2017realvalued}.
%
% Basic components of a GAN
%
\acp{GAN} train two competing neural networks, consisting of a discriminator
network that is set up to distinguish between real and fake data and a
generator network that produces fake versions of the real data. The generator model performs a mapping from a fixed length vector $\mathbf{z}$ to its
representation of the data. The input vector is drawn randomly from a Gaussian distribution which is referred to as a latent space comprised of latent variables. The latent space is a compressed representation of a data distribution which the generator applies meaning to during training. Sampling points from this space allows the generator to produce a variety of different generations, with different points corresponding to different features in the generations. The discriminator maps its input $\mathbf{x}$ to a probability that the input came from either the training (real) data or
generator (fake).
\begin{comment}
Chris' definition of latent space if needed: It maps the intrinsic
variation of the training space onto the Gaussian latent space distribution.
\end{comment}
%
% training a GAN
%
During training, the discriminator and generator are updated using batches of data. Random latent vectors are given to the generator to produce a batch of fake samples and an equal batch of real samples is taken from the training data. The discriminator makes predictions on the real and fake samples and the model is updated through minimising the binary cross-entropy function \cite{Goodfellow-et-al-2016}
%
\begin{equation}
    L = -y \log(\hat{y}) - (1 - y) \log(1-\hat{y}),
    \label{eqn:crossentropy}
\end{equation}
where $\hat{y}$ is the network prediction and $y$ is the true output. While training the discriminator, D, on real data, we set $y = 1$ and $\hat{y} = D(\mathbf{x})$ which from \cref{eqn:crossentropy} gives $L(D(\mathbf{x}),1) = \log(D(\mathbf{x}))$. While training on fake data produced by the generator, G, $y = 0$ and $\hat{y} = D(G(\mathbf{z}))$ and so, $L(D(G(\mathbf{z})),0) = \log(1-(D(G(\mathbf{z}))))$. Since the objective of the discriminator is to correctly classify fake and real data these equations should be maximised, while the goal of the generator should be to minimize these equations. This gives the \ac{GAN} loss as
%
\begin{equation}
   \mathop{\text{min}}_{G}  \mathop{\text{max}}_{D} V(D,G) = \mathbb{E}_{\mathbf{x} \sim p_{\text{\text{data}}}(\mathbf{x})} [\text{log} D(\mathbf{x})] + \mathbb{E}_{\mathbf{z} \sim p_{\mathbf{z}}(\mathbf{z})} [\text{log}(1-D(G(\mathbf{z})))],
 \label{equation:GANloss}
 \end{equation}
where $p_{\text{\text{data}}}(\mathbf{x})$ is the distribution of real data and $p_{\text{z}}(\mathbf{z})$ is the distribution of the latent space.

\subsection{Training stages}
%
Training a GAN is involves updating both the discriminator and generator in stages. First, the discriminator is updated on real instances from the training set. We set the true label $y=1$ and calculate the loss with respect to the predictions $\hat{y}$ via \cref{equation:GANloss}. Stochastic gradient descent is used to minimize the loss which has reduced to $L_D(\textrm{real}) = -\log(\hat{y})$. The discriminator is then trained on fake instances taken from the generator where we set $y=0$ and minimize $L_D(\textrm{fake}) = -\log(1-\hat{y})$. To train the generator, we use a composite model of the generator and discriminator and allow the gradients to flow through this entire model. At this step we set the labels for the generator as real $y=1$ and minimize $L_G = -\log(\hat{y})$. The act of switching the labels and minimizing this form of loss is identical to the previously mentioned generator loss but is beneficial as in practice this function provides more useful gradients. This tweak to the generator loss is called non-saturating generator loss and was reported in the original GAN paper \cite{Goodfellow2014}. It was also shown in that paper that if the generator and discriminator can no longer improve, then the discriminator can no longer distinguish between real and fake i.e. $D(x) = \frac{1}{2}$.

@@

% describe the main idea
%
Given a 100-dimensional vector drawn from a normal distributed latent space and a one-hot encoded class
label, the GAN is able to generate burst-like waveforms generalised from the
training set. We set out by describing the quality of generated waveforms and
how they compare to the training set. We then explore the structure of the
latent and class spaces by interpolating between points in these spaces. We
test two methods of sampling from the class space that can be used to generate
a new signals composed of weighted elements of each training class.
\begin{figure}[!h]
    \centering
    \includegraphics[width=\textwidth]{figures/generations/sg.png}
    \includegraphics[width=\textwidth]{figures/generations/rd.png}
    \includegraphics[width=\textwidth]{figures/generations/wnb.png}
    \includegraphics[width=\textwidth]{figures/generations/blip.png}
    \includegraphics[width=\textwidth]{figures/generations/bbh.png}
    \caption{\ac{GAN} Generated waveforms plotted as a function of time. The latent space inputs for each panel is randomised and each row is assigned one of the five class vectors. By row: sine-Gaussian, Ringdown,
white noise burst, Gaussian pulse, binary black hole merger. On the first four rows each x axis is zoomed for ease of viewing.}
\label{fig:gen_signals} 
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Known class signal generation}
In \cref{fig:gen_signals} we show conditional signal generations using our generator network. We can see the generations are of very good quality and appear like they could plausibly have come from the training set. We can also see that the model has learned the overall characteristics of the five training classes and is able to disentangle each class and associate them with the conditional input. Additionally, as the latent variable changes as do the parameters within each class. For instance \cref{fig:gen_signals} shows how signals vary in frequency, central epoch, decays and phase. The GANs ability to generate a variety of signals for various latent input indicate stable training and no mode collapse. 
\begin{figure}[!h]
    \centering
    \includegraphics[width=\textwidth]{figures/generations/z_interp_sg.png}
    \includegraphics[width=\textwidth]{figures/generations/z_interp_rd.png}
    \includegraphics[width=\textwidth]{figures/generations/z_interp_wnb.png}
    \includegraphics[width=\textwidth]{figures/generations/z_interp_blip.png}
    \includegraphics[width=\textwidth]{figures/generations/z_interp_bbh.png}
    \caption{Generated interpolated waveforms plotted as a function of time. For each interpolation two points were randomly chosen in the latent space and the between panels are linearly interpolated vectors between these two points. Each row keeps on of the class vetors constant throughout the latent space interpolation.}
    \label{fig:z_interp}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Interpolation within the latent space}
%
% expain what we are planning to show here
%
 We have shown that the generator produces quality signals and that the model responds well to randomly sampled Gaussian latent vectors. We now assume that during training the generator has learned a mapping from a Gaussian latent space to signal space and that this mapping is smooth. To test this, we fix the class vector input and linearly interpolate between two randomly chosen point in the latent space. In \cref{fig:z_interp} we show the generated waveforms, with the class vectors held constant along each row. We can see that each plot shows plausible waveforms suggesting that the generator has constructed a smooth traversable, space.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Interpolation between pairs of classes}

@@

\subsection{Results}
In~\cref{fig:eff_curves} we compare the \ac{CNN} results between the three
datasets, we train three \acp{CNN} on the vertex, simplex and uniform data sets
and use these models to make predictions on the other unseen testing data. The top plot is tested on the vertex data and shows that each model confidently classifies all data at 100\% for SNR$>$13. At smaller SNRs the vertex and simplex perform similarly, however, the uniform data performs slightly worse. We would expect that when these models are tested on vertex data that everything should be classified at high SNR as the vertex data is a subset of the uniform and vertex. The middle plot shows the results of the CNNs trained on simplex data.  As expected the simplex and vertex models recover 100\% detection at higher SNRs while the vertex trained model fails this test. This is explained when we consider that the simplex data is a subset of uniform while the vertex data is not. The final plot tests the models on uniform data and show again at high SNRs both simplex and uniform are 100\% detectable as a consequence of our previous subset argument. The vertex test is not fully detectable and further supports our premise that the vertex data, which is based on the modelled training data is distinguishable to both unmodelled datasets.

The three tests shown previously show that the vertex model only manages full detection when tested on itself. The uniform model performs best in all cases and that it contains signals from the vertex and simplex samples, suggesting that this way of sampling the class space as input to the generator creates a good variety of unmodelled generations.

@@

\begin{figure}[h!] % "[t!]" placement specifier just for this example
\centering
\begin{subfigure}{0.7\textwidth}
\includegraphics[width=\linewidth]{figures/efficiency_curve_vtest.png}
\end{subfigure}

\begin{subfigure}{0.7\textwidth}
\centering
\includegraphics[width=\linewidth]{figures/efficiency_curve_stest.png}
\end{subfigure}

\begin{subfigure}{0.7\textwidth}
\centering
\includegraphics[width=\linewidth]{figures/efficiency_curve_utest.png}
\end{subfigure}

\caption{Efficiency curves comparing the performance of the \acp{CNN}. The true alarm probability is plotted as a function of the optimal SNR of the signals. Each plot shows the performance of a CNN trained on vertex, simplex and uniform datasets tested on vertex (top), simplex (middle), uniform (bottom).}
\label{fig:eff_curves}
\end{figure}
